---
title: "hashtag_scRNA"
author: "hui.wan"
date: "12/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, warning = F, message = F)
knitr::opts_knit$set(root.dir = '~/OneDrive - KI.SE/Mac/Project/cowork/20211210_scRNA_hybridvaccine/')
```

```{r load pkgs}
library(Seurat)
library(patchwork)
library(tidyverse)
library(cowplot)
library(Matrix)
library(DoubletFinder)
library(scRepertoire)
library(harmony)
library(clustree)
library(scPred)
library(RColorBrewer)
library(ggpubr)
library(eoffice)

```

# Description

*Aim*:
This experiment seeks to understand the hybrid immunity induced by either Cansino vs. AstraZeneca vaccines. 

*Expermient*:
We have only two reactions. Each reaction has five pools, and each pool has three samples. Samples include patients vaccinated with Cansino or AstraZeneca, with previous or no infection. Each reaction have data for VDJ, GEX, and HashTag.

*Analysis*:
0. Alignment: CellRanger multi (done)
1. Merge data 
2. QC: filter MT
3. Dimensionality reduction, BatchEffector, Clustering
4. Annotation: 
5. scRNAseq: DEGs among vaccine, and different cell types
6. scBCRseq: VDJ recombination. isotype

## set variables
```{r gloable vars}
order_sample = c( "AZ", "AZ-hb", 'Cso', 'Cso-hb', "Inf")
color_sample = structure( brewer.pal(6, 'Paired')[c(1,2,3,4,5)],
                          names = c('Cso', 'Cso-hb', "AZ", "AZ-hb", "Inf"))

color_tf = structure(brewer.pal(9, 'Set1')[c(1,9)], names = c('True', 'False'))

outfile = 'plots/plot_20230401.pptx'
```


# Dimplot

## all cells
```{r all cells}
w = 6
h = 6
reduction = 'umap'

alldata <- read_rds('results/04_qc_dr_int_cl_dge.rds')

# plot sammples
ggsave( DimPlot(alldata, reduction = "umap", group.by = "HTO_classification", repel = T, label = F, label.size = 7) + NoAxes()  +  NoLegend() + 
           scale_color_manual(values = color_sample) +
            theme(panel.background = element_rect(colour = 'black', size = 2))  ,
        filename =  'Figures/dimplot.allcell.HTO.pdf', w = w, h = h)


# plot major cell type
ggsave( DimPlot(alldata, reduction = "umap", group.by = "major_cell_type", repel = T, label = T, label.size = 7) + NoAxes() +  NoLegend()+  
          scale_color_brewer(palette="Set2") + 
          theme(panel.background = element_rect(colour = 'black', size = 2)) ,
        filename =  'Figures/dimplot.major.celltypes.pdf',w = w, h = h)



```


## B cells
```{r all cells}
w = 6
h = 6
reduction = 'umap'



# plot sammples
ggsave( DimPlot(alldata, reduction = "umap", group.by = "HTO_classification", repel = T, label = F, label.size = 7) + NoAxes()  +  NoLegend() + 
           scale_color_manual(values = color_sample) +
            theme(panel.background = element_rect(colour = 'black', size = 2))  ,
        filename =  'Figures/dimplot.allcell.HTO.pdf', w = w, h = h)


# plot major cell type
ggsave( DimPlot(alldata, reduction = "umap", group.by = "major_cell_type", repel = T, label = T, label.size = 7) + NoAxes() +  NoLegend()+  
          scale_color_brewer(palette="Set2") + 
          theme(panel.background = element_rect(colour = 'black', size = 2)) ,
        filename =  'Figures/dimplot.major.celltypes.pdf',w = w, h = h)



```


## B cells
```{r dimplot}
w = 6
h = 6
reduction = 'umap'
#reduction = 'tsne'

q.all.nb <- read_rds('Figures/Bcells.rds')

# plot samples
ggsave( DimPlot(q.all.nb, reduction = reduction, group.by = "HTO_classification") + NoAxes()  +  NoLegend() + 
           scale_color_manual(values = color_sample) + NoLegend() +
          theme(panel.background = element_rect(colour = 'black', size = 2))  ,
        filename =   paste0('Figures/',reduction, '.sample.F.pdf'), w = w, h = h)



# plot cell type
ggsave( DimPlot(q.all.nb, reduction = reduction, group.by = "b.cell.types", repel = T, label = T, label.size = 7) + NoAxes() +  NoLegend()+  
            labs(title = 'Cell identities') + scale_color_brewer(palette="Set2") + theme(panel.background = element_rect(colour = 'black', size = 2)) ,
         filename =   paste0('Figures/',reduction, '.celltypes.F.pdf'), w = w, h = h)
 


#plot isotype
ggsave( DimPlot(q.all.nb, reduction = reduction, group.by = "isotype") + NoAxes()  +  NoLegend()+  
           scale_color_manual(values = col_isotype) + 
          theme(panel.background = element_rect(fill = 'white', colour = 'black', size = 2)),
        filename =   paste0('Figures/',reduction, '.VDJ.isotype.F.pdf'), w = w, h = h)


# plot samples
for (sample in order_sample) {
  q.all.nb@meta.data[sample] <- if_else(q.all.nb$HTO_classification == sample, 'True', "False")

  ggsave( DimPlot(q.all.nb, reduction = reduction, group.by = sample, repel = T, label = F, label.size = 7) + NoAxes()  +  NoLegend() + 
          labs(title = sample ) + 
          #scale_color_manual(values = structure(c(color_sample[sample], '#999999'), names = c('True', 'False' ))) +
          scale_color_manual(values = color_tf) +
          theme(panel.background = element_rect(colour = 'black', size = 2))  ,
        filename =  paste0('Figures/','sample.', sample, '.T.pdf'), w = w, h = h)

}

```



## T cells

```{r dimplot}
w = 6
h = 6

reduction = 'umap'
sel.clust = "RNA_snn_res.2" 

q.all.t.filter <- read_rds('Figures/Tcells.rds')

# plot samples
ggsave( DimPlot(q.all.t.filter, reduction = reduction, group.by = 'HTO_classification') + NoAxes()  +  NoLegend() + 
          scale_color_manual(values = color_sample) +
          theme(panel.background = element_rect(colour = 'black', size = 2))  ,
        filename =  paste0('Figures/',reduction, 'sample.F.pdf'), w = w, h = h)



# plot cell type
ggsave( DimPlot(q.all.t.filter, reduction = reduction, group.by = "t.cell.types", repel = T, label = T, label.size = 7) + NoAxes() +  NoLegend()+  
          #scale_color_brewer(palette="Set2") + 
          theme(panel.background = element_rect(colour = 'black', size = 2)) ,
         filename =   paste0('Figures/',reduction, '.celltypes.F.pdf'), w = w, h = h)
 

# plot for each sample 
for (sample in order_sample) {
  q.all.t.filter@meta.data[sample] <- if_else(q.all.t.filter$HTO_classification == sample, 'True', "False")

  ggsave( DimPlot(q.all.t.filter, reduction = reduction, group.by = sample, repel = T, label = F, label.size = 7) + NoAxes()  +  NoLegend() + 
          labs(title = sample ) + 
          #scale_color_manual(values = structure(c(color_sample[sample], '#999999'), names = c('True', 'False' ))) +
          scale_color_manual(values = color_tf) +
          theme(panel.background = element_rect(colour = 'black', size = 2))  ,
        filename =  paste0('Figures/','sample.', sample, '.T.pdf'), w = w, h = h)

}


```
